---
title: "A flexible approach to time-to-event data analysis using case-base sampling"
author: "Jesse Islam"
date: "July 11, 2019"
output:
  ioslides_presentation:
    fig_caption: yes
    incremental: yes
    keep_md: yes
    widescreen: yes
  slidy_presentation:
    incremental: yes
  beamer_presentation:
    incremental: yes
  bibliography: 010-bibliography.bib
---

```{r packages,echo=FALSE}
knitr::read_chunk("010-packages.R")

```

```{r required-packages, echo=FALSE, warning=FALSE,message=FALSE,results='hide'}
#cleanbib()
options("citation_format" = "pandoc")
bib <- read.bibtex("010-bibliography.bib")
```
--- 

#Motivating Example

## Motivating Example {.emphasized}
* Meet Carol
    * Age: 56
    * Worried about his Prostate
    * What is Carol's two year risk for death by Prostate Cancer?

## Popular methods in time-to-event analysis 

* In disease etiology, we tend to make use of the proportional hazards hypothesis 
    * Cox Regression

* When we want the absolute risk:
    * Parametric models
    * Breslow estimator

# Motivations for a new method
## Motivations for a new method
 *  Julien and Hanley (2008) found that survival analysis rarely produces prognostic functions, even though the software is widely available in cox regression packages.
     * They believe that this is due to the Cumulative incidence curves (or survival curves) being stepwise rather than smooth, reducing interpretability.
 * Parametric model packages don’t usually allow modification of the time component.
 * a streamlined approach for reaching a **smooth absolute risk** curve


#Data on the men in the European Randomized Study of Prostate Cancer Screening (ERSPC)

## ERSPC Data

* ~150 000 men ages 55-69
* First start: 1991
* End: 2006 
*
```{r , echo=FALSE, warning=FALSE,message=FALSE}
ERSPCtable <- flextable( head(ERSPC, n = 7 ), col_keys=c("ScrArm","Follow.Up.Time","DeadOfPrCa"))

ERSPCtable <- bold(ERSPCtable, part = "header")
ERSPCtable <- fontsize(ERSPCtable, size=25)
ERSPCtable <- fontsize(ERSPCtable, part = "header",size=32)
ERSPCtable <- padding( ERSPCtable, padding = 10, part = "all" )
ERSPCtable
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
```


#Casebase: Population-Time plots

##Casebase: Population-Time plots
```{r, results='hide',warning=FALSE}
pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa",time="Follow.Up.Time")
plot(pt_object)
```

#Casebase: Theory

##Casebase: Sampling

* Case-series: all observations where an event occurs
* Base-series: The set of infinite person-moments within the time frame of a study
* Person-moments: a point in time for a given person within the time frame of a study.


##Casebase: Sampling

* c = case-series
* b = base-series
* Should we use a ratio of b=100c, the coefficients are expected to be only 1% larger, using case-base sampling.
* Casebase sampling simplifies the model fitting so that multinomal regression can be used.

$$(\frac{1}{c}+\frac{1}{b})^{-1}$$

##Casebase: GLMs

 $$Let \ odds=e^{L}\\
    \lambda(x,t)*B/b=odds \\
    \lambda(x,t)*B/b=e^{L}\\
    log(\lambda(\hat{x,t}))=log(b/B)+L$$

##Casebase: Likelihood

* For each event type $j=1,...,m$, there exists a non-homogeneous poisson process with a hazard $\lambda_{j}(t)$
* for any given point in time, only one event can occur.
* Censoring is non informative
* casebase sampling also follows a nonhomogeneous Poisson process, where hazard = $\rho(t)$

$$ \prod^{m}_{j=1}\frac{\lambda_{j}(t)^{dN_{j}(t)}}{\rho(t)+\sum^{m}_{j=1}\lambda_{j}(t)}$$

##Casebase: Likelihood

* likelihood is reminiscent to log odds
    * Offset of $log(\rho(t))$ where does this come from??
* Likelihood results in a mean of 0 and is asymptotically normal.
* Gives access to GLM software and statistical tools. (deviance tests, information criteria, regualrization, etc.)

$$ \prod^{m}_{j=1}\frac{\lambda_{j}(t)^{dN_{j}(t)}}{\rho(t)+\sum^{m}_{j=1}\lambda_{j}(t)}$$

```{r , echo=FALSE, results='hide',warning=FALSE}
bibliography()
```
